---
title: "Replacement Class - Introduction to spatial data"
date:  "2023-11-12"
author: "Juliet Thistleton"
output: html_notebook
---

This notebook supports an asynchronous lecture meant to introduce you to some basics of spatial data.  The lecture will be provided via Echo360.

I will provide you with a partially completed notebook to assist your work.

This notebooks will focus on "spatial data" and also accessing demographic data in R.

Key topics for today:

* Using `here` package
* Using the `/data_raw/` directory
* Reminder on joins (focus: left join)
* Our Spatial Data
   * Neighborhoods
   * Joining with non-spatial data
   * Census data
   * Joining with spatial data
* ignore html in git

## Packages

Standards:

```{r}
library(knitr)
library(tidyverse)
library(janitor)
library(lubridate) # because we will probably see some dates
library(here) # more easily access files in your project
```

Some additional packages focuses on today's work:

```{r}
library(sf) # working with simple features - geospatial
library(tmap)
library(tidycensus)

```
A link to a book on `tmap`: https://r-tmap.github.io/


## Using the Neighborhood Geospatial Data (using /data)

I will use the GeoJSON file.  (Newer, not necessarily better, but ... a single file.  Not smaller, but ... this one is not big.)  

https://opendata.dc.gov/datasets/DCGIS::dc-health-planning-neighborhoods/about


## Using the Neighborhood Geospatial Data (using /data)

Load the neighborhood geospatial data as `neigh`.
```{r}
neigh=st_read(here("data_raw", "DC_Health_Planning_Neighborhoods.geojson")) |> clean_names()

class(neigh)
```

Ensure it plots.

```{r}
plot(neigh)
```


## Investigating joining spatial and non-spatial data

Download the DC covid datase for positive cases and store at an appropriate place in your project.

Read the data as `df_c` and be sure to clean names.

```{r}
df_c = read.csv((here("data_raw", "DC_COVID-19_Total_Positive_Cases_by_Neighborhood.csv"))) |> clean_names()
```

Now - let's focus on a particular date (for no reason other than simplifying our analysis).

```{r}
df_cases = df_c |>
  filter(as_date(date_reported) == "2021-11-17") |>
  separate(neighborhood, into = c("code", "name"), sep = ":") |>
  mutate(code = case_when(code == "N35" ~"N0", 
                          TRUE ~ code)) |>
  select(-date_reported)
```


Create the dataframe `df_cases`:

```{r}
df_cases=df_c %>%
  filter(as_date(date_reported) == "2021-11-17") %>% 
  separate(neighborhood,into=c("code","name"),sep = ":") %>%
  mutate(code=case_when(code=="N35" ~"N0",
                        TRUE ~ code)) %>%
  select(-date_reported)
  
  
```

## Regular joining (of dataframes)

Join the dataframes and make a chloropleth map using tmap.

```{r}
neigh2 = left_join(neigh, df_cases, by = c("code")) 

tmap_mode("view")

tm_shape(neigh2) + tm_polygons("total_positives", alpha = .5)

```


## Joining with other spatial data

Let's get some data using `tidycensus`.  Need an API key   https://api.census.gov/data/key_signup.html

```{r}
census_api_key("9fc5d3792d3a5e922287c3f4e9995118766d50a2")

```

 What data is available --- and what is the variable name?
 
 (We are interested in the 5year American Community Survey.)

```{r}
#Data available: Demographics of Dc area
#Variable names: name, label, concept, geography
v20 = load_variables(2018, "acs5")


```

Get some data:

```{r}
df_cencus=get_acs(geography = "tract",
                  variables=c("median_inc"="B06011_001",
                              "pop"="B01001_001",
                              "pop_black"="B02009_001"),
                  state="DC",geometry=TRUE,year=2021) 
```

Make a plot to verify that you read the data:

```{r}
class(df_cencus)
plot(df_cencus)
```


### A BETTER VISUALIZATION

It's in long format.  Let's make it wide.

```{r}
df_cens=df_cencus %>% 
  select(-moe) %>% 
  pivot_wider(names_from = "variable", 
              values_from = "estimate")
  
 

tm_shape(df_cens) +tm_polygons("median_inc",alpha=.5)
```


### How to join

Consider this problem:

```{r}

  tm_shape(neigh2) +tm_borders(col="blue",lwd=5,alpha=.2)+
  tm_shape(df_cens) +tm_borders(col="red",lwd=1,alpha=.3)
```

OK - follow the challenging code elements:

```{r}

#df_j=st_join(df_cens,neigh2)

```

You need to add a coordinate system to the census data:

```{r}
df_cens_adj = df_cens |> st_transform(4326)
```



But which way do we join --- and --- think about how it should "aggregate" the data.

```{r}
df_j = st_join(df_cens_adj, neigh2, largest = TRUE)
```

```{r}
df_j_rev = st_join(neigh2, df_cens_adj, largest = TRUE)
```

Since we want the geometry for the NEIGHBORHOODS, we need a different work a little harder:

```{r}
df1=df_j %>% select(median_inc,pop,pop_black,code) %>%
  group_by(code) %>%
  summarise(pop_n=sum(pop),
            pop_black_n=sum(pop_black), 
            adj_median_income=sum(pop*median_inc)/pop_n) 

plot(df1)
```


Now that we are aggregating in the right way, we can join.

```{r}
#df2=left_join(neigh2,df1)

df2=left_join(neigh2,df1 %>% st_set_geometry(NULL))

```

And visualize:

```{r}
df2=df2 %>% 
  mutate(black_perc=pop_black_n/pop_n, covid_rate=total_positives/pop_n)
tm_shape(df2)+tm_polygons(c("adj_median_income","covid_rate","black_perc"))
```

Improve that visualization:

```{r}
df2 |> 
    mutate(black_perc=pop_black_n/pop_n, covid_rate=total_positives/pop_n) |>
  filter(code != "N0") |>
  tm_shape() + tm_polygons(c("adj_median_income", "covid_rate", "black_perc"), alpha = .4)
```
